Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-08-04 psy_peds.Rmd
after the pandemic changed into during the pandermic; first graphic sorted in descending order

We will also transform the observation dates, and break it by weeks, adding 2 additional columns: "weeks" and "year". Finally, we will dichotomize the study timeframe in two periods, based on cut-off dates, "before_pandemic", and "after_pandemic".

@@

### Data management for the patientClinicalCourse table
```{r}
clinical_raw <- clinical_raw %>% 
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks= "week")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic","after_pandemic")) %>% 
  dplyr::filter(cohort  %in% c("AllAdm", "AllNotAdm"))
```

@@

## Patients having presented history of mental health related issues before pandemic
We estimate the number of patients that presented a mental health disorder (grouped by mental health disorder groups - icdCodes$disorder_group):
- only before the pandemic
- only after the pandemic
- both before and after pandemic

@@

```{r}
patient_count_barplot_period <- input_data %>% 
  dplyr::group_by( disorder_group, patient_num, cohort) %>% 
  dplyr::mutate( 
    when = ifelse( "before_pandemic" %in% period & "after_pandemic" %in% period,
                   "Before and After Pandemic",
                   ifelse( "before_pandemic" %in% period & ! "after_pandemic" %in% period,
                           "Only Before Pandemic",
                           "Only After Pandemic"
                   )
    ) 
    %>% factor(levels = c("Only Before Pandemic", "Before and After Pandemic", "Only After Pandemic"))
  ) %>%
  dplyr::ungroup() %>% 
  dplyr::select( patient_num, cohort, disorder_group, when ) %>%
  dplyr::group_by( disorder_group, when ) %>%
  dplyr::summarise( count = n_distinct( patient_num, cohort ) ) %>% 
  dplyr::mutate(count = ifelse(count > obfuscation | isFALSE(obfuscation),
                               count, 
                               0.5), 
                siteid = site)

patient_count_barplot_period %>%
  ggplot(aes(x = as.factor( disorder_group), y = count, fill = when)) +
  geom_bar(stat="identity", position = "dodge") +
  scale_fill_manual(values = cbPalette, name = "Period", guide = guide_legend(reverse = TRUE)) +
  labs(y = "Unique patients count",
       x = "Disorder group",
       title = "Patients with at least one visit", 
       subtitle = "According to visit period and pathology group", 
       legend = "Period")+
  theme(legend.position = "bottom", 
        legend.direction = "vertical",
        title = element_text(size=10),
        plot.subtitle = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text.x = element_text(size =7, angle = 0, hjust = 0.5, vjust = 0)) + 
  coord_flip()

ggsave("output/disorder_before_after_pandemic_all.png", width = 18, height = 9, units="cm")
```

@@

#estimate the number of patients per week
patient_count_per_week %>%
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "after_pandemic")) %>% 
  ggplot(aes(x = weeks, y = n, fill = period)) +
  geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

obs_by_day <- encounters_obs %>% 
  filter(concept_code %in% icdCodes$ICD10_Code) %>% 
  left_join(clinical_raw[clinical_raw$in_hospital == 1, c("patient_num", "cohort", "n_hospitalisation", "calendar_date")], 
            on = c("patient_num", "n_hospitalisation")) %>% 
  group_by(calendar_date) %>% 
  summarise(patient_count = n_distinct(patient_num, cohort)) %>% 
  mutate(calendar_date = as.Date(calendar_date, format=dateFormat), 
         period = if_else(calendar_date < pandemic_start_date,
                          "Before Pandemic",
                          "After Pandemic"
         ), 
         patient_count = ifelse(patient_count > obfuscation | isFALSE(obfuscation), 
                                 patient_count, 
                                 0.5)
  )

@@

patient_count_per_week_disorder_group %>%
  mutate(weeks = as.Date(weeks), 
         period = if_else(weeks < pandemic_start_date,
                          "Before Pandemic",
                          "After Pandemic"
         )
  ) %>% 
  ggplot(aes(x = as.Date(weeks), y= n, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  #geom_vline(xintercept = as.Date(late_pandemic_start_date),
  #           linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Week",
       y = "Patient Counts",
       title = "Weekly patient counts with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])
ggsave("output/categorical_counts_inpatient_teens_all.png", width = 18, height = 9, units="cm")
```

@@

summary_race_test <- race_by_period %>% 
  select(-percentage) %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  select(after_pandemic) %>% 
  tidyr::replace_na(list(after_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")
}
```

@@

summary_diagnosis_test <- summary_diagnosis %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  #select(late_pandemic, early_pandemic) %>% 
  select(after_pandemic) %>%
  #tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>% 
  tidyr::replace_na(list(after_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")

@@

summary_disorder_group_by_period_test <- summary_disorder_group_by_period %>% 
  pivot_wider(id_cols = disorder_group, names_from = period, values_from = c(count, percentage)) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, replace = 0))) %>% 
  #select(count_early_pandemic, count_late_pandemic) %>% 
  select(count_after_pandemic) %>%
  chisq.test() %>% 
  `[[`("p.value")

@@

diagnosis_table1 <- summary_diagnosis %>% 
    bind_rows(
    sex_by_period %>%
      group_by(period) %>% 
      dplyr::summarise(concept_code = "Total", count = sum(count))
  ) %>%
  pivot_wider( names_from = period, values_from = count) %>% 
  mutate(diag = paste(concept_code, description, sep = ": ")) %>% 
  select(diag, after_pandemic) %>% 
  rename(name = diag)

disorder_table1 <- summary_disorder_group_by_period %>% 
  bind_rows(
    dplyr::summarise(., disorder_group="Total", across(where(is.numeric), sum))
  ) %>% 
  mutate(result = format_counts(count, percentage)) %>% 
  select(-c(count, percentage)) %>% 
  pivot_wider( names_from = period, values_from = result ) %>% 
  # replace_na( list(early_pandemic = "0 (0 %)", 
  #                  late_pandemic = "0 (0 %)")) %>% 
    replace_na( list(after_pandemic = "0 (0 %)")) %>% 
  rename(name = disorder_group)

@@

table1 %>% 
  group_by(Category) %>%
  mutate(pvalue = ifelse(duplicated(pvalue, fromLast = TRUE), "", as.character(pvalue))) %>%
  ungroup() %>%
  mutate(Category = "") %>%
  mutate(across(.cols = c(before_pandemic, after_pandemic), .fns = replace_na, "0")) %>% 
  rename_with(str_to_title) %>%
  rename_with(sub, pattern = "_", replacement = " ") %>%
  mutate(across(.fns = str_to_title)) %>% 
  mutate(across(.fns = sub, pattern = "_", replacement = " ")) %>% 
  kable( booktabs = T) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::group_rows(index = setNames(rle(table1$Category)[[1]], rle(table1$Category)[[2]]))
```

>>>>

We will also transform the observation dates, and break it by weeks, adding 2 additional columns: "weeks" and "year". Finally, we will dichotomize the study timeframe in two periods, based on cut-off dates, "before_pandemic", and "during_pandemic".

@@

### Data management for the patientClinicalCourse table
```{r}
clinical_raw <- clinical_raw %>% 
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks= "week")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic","during_pandemic")) %>% 
  dplyr::filter(cohort  %in% c("AllAdm", "AllNotAdm"))
```

@@

## Patients having presented history of mental health related issues before pandemic
We estimate the number of patients that presented a mental health disorder (grouped by mental health disorder groups - icdCodes$disorder_group):
- only before the pandemic
- only during the pandemic
- both before and during pandemic

@@

```{r}
patient_count_barplot_period <- input_data %>% 
  dplyr::group_by( disorder_group, patient_num, cohort) %>% 
  dplyr::mutate( 
    when = ifelse( "before_pandemic" %in% period & "during_pandemic" %in% period,
                   "Before and During Pandemic",
                   ifelse( "before_pandemic" %in% period & ! "during_pandemic" %in% period,
                           "Before the Pandemic",
                           "During the Pandemic"
                   )
    ) 
    %>% factor(levels = c("Before the Pandemic", "Before and During Pandemic", "During the Pandemic"))
  ) %>%
  dplyr::ungroup() %>% 
  dplyr::select( patient_num, cohort, disorder_group, when ) %>%
  dplyr::group_by( disorder_group, when ) %>%
  dplyr::summarise( count = n_distinct( patient_num, cohort ) ) %>% 
  dplyr::mutate(count = ifelse(count > obfuscation | isFALSE(obfuscation),
                               count, 
                               0.5), 
                siteid = site)

# create a vector with the disorder groups sorted in decreasing order 
order_disorders <- patient_count_barplot_period %>% 
  filter( when == "During the Pandemic") %>%
  arrange( count ) %>%
  select( disorder_group )

patient_count_barplot_period$orderDisorder <- factor(patient_count_barplot_period$disorder_group,levels = as.factor( order_disorders$disorder_group ))

patient_count_barplot_period %>%
  ggplot(aes(x = as.factor( orderDisorder), y = count, fill = when)) +
  geom_bar(stat="identity", position = "dodge") +
  scale_fill_manual(values = cbPalette, name = "Period", guide = guide_legend(reverse = TRUE)) +
  labs(y = "Unique patients count",
       x = "Disorder group",
       title = "Patients with at least one visit", 
       subtitle = "According to visit period and pathology group", 
       legend = "Period")+
  theme(legend.position = "bottom", 
        legend.direction = "vertical",
        title = element_text(size=10),
        plot.subtitle = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.text.x = element_text(size =7, angle = 0, hjust = 0.5, vjust = 0)) + 
  coord_flip()

ggsave("output/disorder_before_during_pandemic_all.png", width = 18, height = 9, units="cm")
```

### Relative to the total number of patients in each period

```{r}
patients_per_period <- input_data %>%
  dplyr::group_by( period ) %>%
  dplyr::summarise( count = n_distinct( patient_num ) ) %>% 
  dplyr::mutate(count = ifelse(count > obfuscation | isFALSE(obfuscation),
                               count, 
                               0.5), 
                siteid = site)

@@

#estimate the number of patients per week
patient_count_per_week %>%
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic")) %>% 
  ggplot(aes(x = weeks, y = n, fill = period)) +
  geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

obs_by_day <- encounters_obs %>% 
  filter(concept_code %in% icdCodes$ICD10_Code) %>% 
  left_join(clinical_raw[clinical_raw$in_hospital == 1, c("patient_num", "cohort", "n_hospitalisation", "calendar_date")], 
            on = c("patient_num", "n_hospitalisation")) %>% 
  group_by(calendar_date) %>% 
  summarise(patient_count = n_distinct(patient_num, cohort)) %>% 
  mutate(calendar_date = as.Date(calendar_date, format=dateFormat), 
         period = if_else(calendar_date < pandemic_start_date,
                          "Before the Start of the Pandemic",
                          "After the Start of the Pandemic"
         ), 
         patient_count = ifelse(patient_count > obfuscation | isFALSE(obfuscation), 
                                 patient_count, 
                                 0.5)
  )

@@

patient_count_per_week_disorder_group %>%
  mutate(weeks = as.Date(weeks), 
         period = if_else(weeks < pandemic_start_date,
                          "Before the Start of the Pandemic",
                          "After the Start of the Pandemic"
         )
  ) %>% 
  ggplot(aes(x = as.Date(weeks), y= n, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  #geom_vline(xintercept = as.Date(late_pandemic_start_date),
  #           linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Week",
       y = "Patient Counts",
       title = "Weekly patient counts with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])
ggsave("output/categorical_counts_inpatient_teens_all.png", width = 18, height = 9, units="cm")
```

@@

summary_race_test <- race_by_period %>% 
  select(-percentage) %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  select(during_pandemic) %>% 
  tidyr::replace_na(list(during_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")
}
```

@@

summary_diagnosis_test <- summary_diagnosis %>% 
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  #select(late_pandemic, early_pandemic) %>% 
  select(during_pandemic) %>%
  #tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>% 
  tidyr::replace_na(list(during_pandemic = 0)) %>% 
  chisq.test() %>% 
  `[[`("p.value")

@@

summary_disorder_group_by_period_test <- summary_disorder_group_by_period %>% 
  pivot_wider(id_cols = disorder_group, names_from = period, values_from = c(count, percentage)) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, replace = 0))) %>% 
  #select(count_early_pandemic, count_late_pandemic) %>% 
  select(count_during_pandemic) %>%
  chisq.test() %>% 
  `[[`("p.value")

@@

diagnosis_table1 <- summary_diagnosis %>% 
    bind_rows(
    sex_by_period %>%
      group_by(period) %>% 
      dplyr::summarise(concept_code = "Total", count = sum(count))
  ) %>%
  pivot_wider( names_from = period, values_from = count) %>% 
  mutate(diag = paste(concept_code, description, sep = ": ")) %>% 
  select(diag, during_pandemic) %>% 
  rename(name = diag)

disorder_table1 <- summary_disorder_group_by_period %>% 
  bind_rows(
    dplyr::summarise(., disorder_group="Total", across(where(is.numeric), sum))
  ) %>% 
  mutate(result = format_counts(count, percentage)) %>% 
  select(-c(count, percentage)) %>% 
  pivot_wider( names_from = period, values_from = result ) %>% 
  # replace_na( list(early_pandemic = "0 (0 %)", 
  #                  late_pandemic = "0 (0 %)")) %>% 
    replace_na( list(during_pandemic = "0 (0 %)")) %>% 
  rename(name = disorder_group)

@@

table1 %>% 
  group_by(Category) %>%
  mutate(pvalue = ifelse(duplicated(pvalue, fromLast = TRUE), "", as.character(pvalue))) %>%
  ungroup() %>%
  mutate(Category = "") %>%
  mutate(across(.cols = c(before_pandemic, during_pandemic), .fns = replace_na, "0")) %>% 
  rename_with(str_to_title) %>%
  rename_with(sub, pattern = "_", replacement = " ") %>%
  mutate(across(.fns = str_to_title)) %>% 
  mutate(across(.fns = sub, pattern = "_", replacement = " ")) %>% 
  kable( booktabs = T) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::group_rows(index = setNames(rle(table1$Category)[[1]], rle(table1$Category)[[2]]))
```
